Museum specimens as baseline for assessing a potential global pollinator crisis

### 2018-03-27 ToDo.Rmd
Make quick notes on whats left to do.
integrated comments/changes from Darren and myself. Have added segments to section on global pollinaotr declines (what we aready know)
Add reference notes to TODO

- check formating rules (words in abstract, main body, etc...)
- correct fig1 for country size?
- Species declines in % for different taxa... what to do?
- think on gbif plot...
- Spain.

@@

@Ollerton2012 
@Winfree2011 - >

>>>>

- check formating rules (words in abstract, main body, etc...)
- correct fig1 for country size?
- Species declines in % for different taxa... what to do?
- for NZ species abundance analysis - should data be split by exotic/native first?
- think on gbif plot...
- Spain.
- Refs

@@

@Ollerton2012 -> Ollerton winfree and tarrant Oikos, I think 2012.
@Buchman1996 -> http://www.worldcat.org/title/forgotten-pollinators/oclc/34077459
@NAS -> https://www.nap.edu/catalog/11761/status-of-pollinators-in-north-america
@Winfree2011 -> Winfree bartomeus Cariveau ARES
@PickeringAscher -> Ascher J.S., Pickering J. (2015) Discover Life bee species guide and world checklist (Hymenoptera: Apoidea: Anthophila). www.discoverlife.org
@Matrin et al2012: https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/110154?journalCode=fron
@Archer: https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1600-0706.2013.00949.x
@França: França, Filipe, et al. "Do space‐for‐time assessments underestimate the impacts of logging on tropical biodiversity? An Amazonian case study using dung beetles." Journal of Applied Ecology 53.4 (2016): 1098-1105.
@DePalma2018: De Palma, Adriana, et al. "Challenges With Inferring How Land-Use Affects Terrestrial Biodiversity: Study Design, Time, Space and Synthesis." Next Generation Biomonitoring 58 (2018): 163.
@DePalma2017 De Palma, Adriana, et al. "Dimensions of biodiversity loss: Spatial mismatch in land‐use impacts on species, functional and phylogenetic diversity of European bees." Diversity and Distributions 23.12 (2017): 1435-1446.
@DePalma2016: De Palma, Adriana, et al. "Predicting bee community responses to land-use changes: Effects of geographic and taxonomic biases." Scientific Reports 6 (2016): 31153.
@vanSwaay: van Swaay, Chris AM, et al. "Butterfly monitoring in Europe: methods, applications and perspectives." Biodiversity and Conservation 17.14 (2008): 3455-3469.
@Scheper2014. PNAS Museum specimens reveal loss of pollen host plants as key factor driving wild bee decline in The Netherlands
@young Are pollinating hawk moths declining in the Northeastern United States? An analysis of collection records
@helminth helminthR: an R interface to the London Natural History Museum's Host–Parasite Database
@keil Biodiversity change is scale‐dependent: an example from Dutch and UK hoverflies (Diptera, Syrphidae)
P Keil, JC Biesmeijer, A Barendregt, M Reemer, WE Kunin
Ecography 34 (3), 392-401
@@taxonomy papers (2):
https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1523-1739.2004.00233.x
Huber, J. T. "The importance of voucher specimens, with practical guidelines for preserving specimens of the major invertebrate phyla for identification." Journal of Natural History 32.3 (1998): 367-385.
@ref geonames, geoparser and opencage: thos are likns to software.
@Hart Ten simple rules for digital data storage. DOI: 10.7287/peerj.preprints.1448v1
@collado2018: https://www.biorxiv.org/content/early/2018/03/08/278812
@Bartomeus_Winfree2013 F1000 https://f1000research.com/articles/2-146/v1
@Aizen Morales, Carolina L., et al. "Rapid ecological replacement of a native bumble bee by invasive species." Frontiers in Ecology and the Environment 11.10 (2013): 529-534.
@Pauw: Reconstruction of historical pollination rates reveals linked declines of pollinators and plants
A Pauw, JA Hawkins - Oikos, 2011
Cited by 53 Related articles All 11 versions
@Bartomeus2011 PNAS phenology
@Tylianakis2010ELE Ecology letters It's 2008 sorry
@Kleijn raemakers: https://www.ncbi.nlm.nih.gov/pubmed/18705369
@refAlaskaScience http://www.sciencemag.org/content/349/6255/1541
@Oliveira http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0148983
@R just cite R
@rgbif https://cran.r-project.org/web/packages/rgbif/rgbif.pdf
@cameron2013] PNAS paper on bumblebees
@OllertonScience Science paper on uk bee extinctions

#to add

@Mikheyev Museum samples reveal rapid evolution by wild honey bees exposed to a novel parasite

### 2018-03-27 ms.Rmd
Add reference notes to TODO

Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [# ???] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@ref]. However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@ref]. Since then, studies on pollinator responses to various global change drivers have multiplied rapidly. Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @ Potts2010]. Yet, the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species check number fell into the “Data Deficient” category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011]. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017]. However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification. Furthermore, the uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos]. Another barrier to assessing pollinator declines is that pollinators are often highly mobile, short-lived and small.

One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@ref], this is the most extensive source of data on pollinator populations currently available. In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [DePalma2017]. A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa [@ref]. However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations. Finally, the most practical approach for assessing long-term pollinator population trends is to use museum specimen collection data.

The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification [@primak]. Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization. Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014). This is despite the fact that museum collections often have a number of biases including: the unknown sampling effort that was employed; the personal interests and curatorial techniques of collectors (e.g. discarding damaged individuals, only accessioning a certain number of individuals, targeting rare or unusual over common taxa); the spatial biases where areas have been undersampled, or where samples are biased towards easily collected localities (e.g. near towns/cities and/or along roadsides);that information is often restricted only to the presence of a species (i.e. there is no information on where a species is absent);and the difficulty of getting information on other taxa from the same location (e.g. NHCs are organised taxonomically, not geographically [#@Keil; other @refs?]. Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.

@@

Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@ref cory paper on taxonomy needs], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@ref on database management]. Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution photos of specimens and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers. In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage]. [Explain zooniverse, NZ initiative and how data can be validated… Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]

@@

Outside of Europe and the US there are no studies on pollinator declines using historical museum collection records. South Africa: Pauw, maybe with plants… . However, there are species-specific examples of historical (e.g., Bombus dalbhomi; #@Aizen).
add % in text also
Justify focus on bees? We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes to highlight the little coverage that we have globally.

@@

GBIF [#@ref] is a central repository centralizing occurrence data from around the globe. Much of this data comes from museums and taxonomic collections [from Darren: Wider than just a ‘museum’ in the traditional sense; also a lot goes on at govt research institutes that have a collection (eg like NZAC), and also sometimes at Universities,], but it integrates several other sources. Alongside the popular statistical language R [#@ref], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest. In Figure 2 we show the number of modern and historical bee records already available per country. While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes. Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis. On the contrary, for countries with large number of historical records, re-surveys should be organized [e.g. #@cameron2013]. Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.

@@

As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors. Using two case studies, we show that collaboration with both museum curators and ecologists is key to adequately understanding the data and treating it appropriately. We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share. [from Darren: I don’t think this is a good statement. It can be deleted and the discussion is still good. Saying “instead of a central museum..” is an argument against the main part of this paper (early in introduction) and also is against the discussion about collaboration]. Ultimately, centralization of regional and national museum data in existing global platforms such as GBIF would facilitate free and widespread access. However, ecological [from Darren: Adding “ecological” means the data is different from the museum records in a central repository. Ecological datasets also do not fit well into museums databases unless there are voucher specimens from the ecology project] datasets can also be stored in different webpages or database repositories (e.g. Dryad), and can be retrieved and combined with other datasets using open science tools (#@ropensci), providing these tools are accessible [API e.g., # @heltmint].

@@

Ultimately, to identify global trends in pollinator declines we require robust data, collected from diverse geographic regions. Further, it is crucial that data are analysed appropriately. This requires researches to identify biases in the data and fill gaps in knowledge of pollinators [gaps in what? taxonomic, geographic? habitat?] where possible. We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees. For the US and Europe there have been few regional bee extinctions [#@Ollerton; Bartomeus2013c] (#check this Bartomeus reference), but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006]. For most other pollinator taxa and regions throughout the world we know almost nothing. Moving forward, the first step for many taxa will be to identify and describe species. Only then can we begin to document declines.

>>>>

Animal pollinators are a critical component of both natural and agricultural ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. The vulnerability of pollinators was identified several decades ago, and was popularized in 1996 by the influential book “the forgotten pollinators” [@Buchman1996]. However, early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to assess the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that “For most pollinator species […] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult” [@nas]. Since then, studies on pollinator responses to various global change drivers have multiplied rapidly. Researchers have now developed strong consensus that disturbances such as habitat destruction, land-use intensification, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010]. Yet, the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species check number fell into the “Data Deficient” category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and include bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011]. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017]. However, bees are also diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and often require expert taxonomists for identification. Furthermore, the uneven distribution of pollinator researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are taxonomic biases toward species that are easier to identify, such as bumblebees [#@ArcherOikos, @DePalma2016]. Another barrier to assessing pollinator declines is that pollinators are often highly mobile, short-lived and small.

One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@Franca, @DePalma2018], this is the most extensive source of data on pollinator populations currently available. In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models using pollinator associations to different land uses [@DePalma2017]. A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs [@vanSwaay] and is currently being extended to other pollinator taxa. However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations. Finally, the most practical approach for assessing long-term pollinator population trends is to use museum specimen collection data.

The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens, and at the same time often ensure higher quality taxonomic identification. Yet, the major bottleneck for researchers wanting to use these data is the lack of digitization. Where digitization has been completed, the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., [@Bartomeus2013a]; # @Scheper2014, @Young). This is despite the fact that museum collections often have a number of biases including: the unknown sampling effort that was employed; the personal interests and curatorial techniques of collectors (e.g. discarding damaged individuals, only accessioning a certain number of individuals, targeting rare or unusual over common taxa); the spatial biases where areas have been undersampled, or where samples are biased towards easily collected localities (e.g. near towns/cities and/or along roadsides);that information is often restricted only to the presence of a species (i.e. there is no information on where a species is absent);and the difficulty of getting information on other taxa from the same location (e.g. NHCs are organised taxonomically, not geographically [#@Keil; other @refs?]. Nevertheless, in many cases, museum collections are the only data we have to reconstruct the past.

@@

Digitizing old collection specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@taxonomy], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@hart]. Undertaking this process for tens or hundreds of thousands of museum collections specimens can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution photos of specimens and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers. In addition, new algorithms have been created that allow location geo-referencing based on vernacular names [@ref geonames, geoparser and opencage]. [Explain zooniverse, NZ initiative and how data can be validated… Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]

@@

Outside of Europe and the US there are no studies on pollinator declines using historical museum collection records. South Africa: @Pauw, maybe with plants… . However, there are species-specific examples of historical (e.g., Bombus dalbhomi; #@Aizen).
add % in text also
Justify focus on bees? We compiled all this information to show several worldwide maps showing knowledge about bee richness (a), Bee species richness changes (b), syrphid species richness changes (c) and butterfly species richness changes to highlight the little coverage that we have globally.

@@

GBIF (www.gbig.com?) is a central repository centralizing occurrence data from around the globe. Much of this data comes from museums and taxonomic collections [from Darren: Wider than just a ‘museum’ in the traditional sense; also a lot goes on at govt research institutes that have a collection (eg like NZAC), and also sometimes at Universities,], but it integrates several other sources. Alongside the popular statistical language R [#@R], GBIF can be directly queried into your computer [#@rgbif] and data availability checked for your region of interest. In Figure 2 we show the number of modern and historical bee records already available per country. While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes. Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis. On the contrary, for countries with large number of historical records, re-surveys should be organized [e.g. #@cameron2013]. Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X… Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.

@@

As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors. Using two case studies, we show that collaboration with both museum curators and ecologists is key to adequately understanding the data and treating it appropriately. We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share. [from Darren: I don’t think this is a good statement. It can be deleted and the discussion is still good. Saying “instead of a central museum..” is an argument against the main part of this paper (early in introduction) and also is against the discussion about collaboration]. Ultimately, centralization of regional and national museum data in existing global platforms such as GBIF would facilitate free and widespread access. However, ecological [from Darren: Adding “ecological” means the data is different from the museum records in a central repository. Ecological datasets also do not fit well into museums databases unless there are voucher specimens from the ecology project] datasets can also be stored in different webpages or database repositories (e.g. Dryad), and can be retrieved and combined with other datasets using open science tools (ropensci), providing these tools are accessible [API e.g., # @heltmint].

@@

Ultimately, to identify global trends in pollinator declines we require robust data, collected from diverse geographic regions. Further, it is crucial that data are analysed appropriately. This requires researches to identify biases in the data and fill gaps in knowledge of pollinators [gaps in what? taxonomic, geographic? habitat?] where possible. We need to place increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees. For the US and Europe there have been few regional bee extinctions [#@OllertonScience; Bartomeus2013c] (#Ref bartomeus should be PNAS historical declines), but in disturbed ecosystems declines are widespread [@DePalma2017; @Bartomeus2013a; @Biesmeijer2006]. For most other pollinator taxa and regions throughout the world we know almost nothing. Moving forward, the first step for many taxa will be to identify and describe species. Only then can we begin to document declines.
